---
title: "r-project"
author: "Riley Maher, Griffin Rubin, Katie Summers, Natalie Vadasz"
date: "Dec 8, 2020"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Group Project

## Import Packages

```{r}
library("ggplot2")
library('dplyr')
library('tidyverse')
library('geosphere')
library("ggmap")
```

## Importing Data

```{r}
# Reading in the sample CSV of rider data we made
rider_2019_sample <- read.csv('sample.csv', stringsAsFactors = TRUE)
head(rider_2019_sample)

# Reading in the weather data set
weather_data <- read.csv('NYCWeather2019.csv', stringsAsFactors = TRUE)
head(weather_data)
```

## Initial Data Summary

```{r}
# Initial summary of rider data set
str(rider_2019_sample)
summary(rider_2019_sample)
```

```{r}
# Initial summart of weather data set
str(weather_data)
summary(rider_2019_sample)
```

### Adjusting Dates in Data Sets

```{r}
# Creating columns of just month, day, and year
weather_data$DATE <- as.Date(weather_data$DATE, format = "%m/%d/%Y")
weather_data$Month <- format(weather_data$DATE, "%m")
weather_data$Day <- format(weather_data$DATE, "%d")
weather_data$Year <- format(weather_data$DATE, "%Y")
```

```{r}
# Creating columns of just month, day, and year
rider_2019_sample$DATE <- as.Date(rider_2019_sample$starttime, format = "%Y-%m-%d")
rider_2019_sample$Month <- format(rider_2019_sample$DATE, "%m")
rider_2019_sample$Day <- format(rider_2019_sample$DATE, "%d")
rider_2019_sample$Year <- format(rider_2019_sample$DATE, "%Y")
```

### Rider Age

```{r}
rider_2019_sample$age <- 2019 - as.numeric(as.character(rider_2019_sample$birth.year))
rider_2019_sample <- filter(rider_2019_sample, age <= 80)
```

### Combining Data Sets

```{r}
# Combining data frames to compare data
edited_weather <- select(weather_data,
                         PRCP,
                         SNOW,
                         AWND,
                         DATE)
edited_rider <- select(rider_2019_sample, 
                       age,
                       gender,
                       usertype,
                       tripduration,
                       start.station.latitude,
                       start.station.longitude,
                       start.station.id,
                       start.station.name,
                       end.station.latitude,
                       end.station.longitude,
                       end.station.id,
                       end.station.name,
                       DATE,
                       Day,
                       Month,
                       Year)

total_data = merge(edited_weather, edited_rider, by.x="DATE", by.y="DATE", all.x=TRUE)
head(total_data)
```

### Distance Between Stations

```{r}
# Distance between start and end station in Meters
total_data <- mutate(total_data, 
                            distance = distHaversine(cbind(total_data$start.station.longitude,
                                                           total_data$start.station.latitude),
                                                     cbind(total_data$end.station.longitude,
                                                           total_data$end.station.latitude)))

head(total_data)
```

## Initial Data Analysis

### Gender Split in Riders

```{r}
# Reclassifying the genders
# 0=unknown, 1=male, 2=female
total_data$gender <- ifelse(total_data$gender == 0, "Unknown",
                                  ifelse(total_data$gender == 1, "Male", "Female"))

# Seeing the split of genders who rented bikes
total_data %>%
  ggplot(aes(x=gender)) +
  geom_bar()
```

### Subscriber vs Customer for Riders

```{r}
# Seeing the split of user type who rented bikes
total_data %>%
  ggplot(aes(x=usertype)) +
  geom_bar()
```

### Trip Duration

```{r}
# Range of all bike rides
total_data <- filter(total_data, tripduration <= 3600)
duration_range <- range(total_data$tripduration, na.rm=TRUE)
duration_range

# Average length of a bike ride
duration_mean <- mean(total_data$tripduration, na.rm=TRUE)
duration_mean

# Standard deviation of bike rides
duration_sd <- sd(total_data$tripduration, na.rm=TRUE)
duration_sd
```
### Trip Duration with Rain

```{r}
# Range of all bike rides affected by rain
total_data_rain <- filter(total_data, SNOW == 0, PRCP > 0)
duration_range_rain <- range(total_data_rain$tripduration, na.rm=TRUE)
duration_range_rain

# Average length of a bike ride affected by rain
duration_mean_rain <- mean(total_data_rain$tripduration, na.rm=TRUE)
duration_mean_rain

# Standard deviation of bike rides affected by rain
duration_sd_rain <- sd(total_data_rain$tripduration, na.rm=TRUE)
duration_sd_rain
```

### Trip Duration with Snow

```{r}
# Range of all bike rides affected by snow
total_data_snow <- filter(total_data, SNOW > 0)
duration_range_snow <- range(total_data_snow$tripduration, na.rm=TRUE)
duration_range_snow

# Average length of a bike ride affected by snow
duration_mean_snow <- mean(total_data_snow$tripduration, na.rm=TRUE)
duration_mean_snow

# Standard deviation of bike rides affected by snow
duration_sd_snow <- sd(total_data_snow$tripduration, na.rm=TRUE)
duration_sd_snow
```

### Trip Duration with Wind

```{r}
# Range of all bike rides affected by wind
total_data_wind <- filter(total_data, SNOW == 0, PRCP == 0, AWND > 0)
duration_range_wind <- range(total_data_wind$tripduration, na.rm=TRUE)
duration_range_wind

# Average length of a bike ride affected by wind
duration_mean_wind <- mean(total_data_wind$tripduration, na.rm=TRUE)
duration_mean_wind

# Standard deviation of bike rides affected by wind
duration_sd_wind <- sd(total_data_wind$tripduration, na.rm=TRUE)
duration_sd_wind
```

### Types of Weather per Month

```{r}
# Average rain per month
total_data %>%
  filter(SNOW == 0) %>%
  summarise(average_rain = tapply(PRCP, Month, mean, na.rm=TRUE))
```

```{r}
# Average snow per month
total_data %>% 
  summarise(avg_snow = tapply(SNOW, Month, mean, na.rm=TRUE))
```

```{r}
# Average wind speed per month
total_data %>%
  summarise(average_wind_speed = tapply(AWND, Month, mean, na.rm=TRUE))
```

## Exploratory Data Analysis - Weather Effects

### Average Rain by Age

```{r}
# Trip duration by age of riders and rain amount
plot_data <- total_data %>%
  filter(SNOW == 0) %>%
  group_by(age) %>%
  summarise(mean_PRCP_by_age = mean(PRCP),
            mean_duration = mean(tripduration)) 

plot_data %>%
  ggplot(aes(x = age, y = mean_PRCP_by_age)) +
  geom_point(alpha =0.9, shape = 18, colour = "blue", size = plot_data$mean_duration/150) +
  geom_smooth(colour = "orange") 
```

### Average Wind by Age

```{r}
# Mean Wind by Age of Rider
total_data %>% 
  group_by(age) %>%
  summarise(mean_AWND_by_age = mean(AWND,na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = mean_AWND_by_age)) + geom_line() + geom_smooth() 
```

## Rain Effects on Trip Duration

```{r}
# Average ride time when it's raining
total_data %>%
  filter(PRCP > 0, SNOW == 0) %>%
  summarise(prcp_duration_mean = mean(tripduration))

total_data %>% 
  filter(PRCP > 0, SNOW == 0) %>%
  ggplot(aes(x = tripduration)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") +
  geom_density(alpha=.2, fill="#FF6666") + 
  geom_vline(aes(xintercept=mean(tripduration)), color="blue", linetype="dashed", size=1)
```

### Snow Effects on Trip Duration

```{r}
# Average ride time when it's snowing
total_data %>%
  filter(SNOW > 0) %>%
  summarise(snow_duration_mean = mean(tripduration))

total_data %>% 
  filter(SNOW > 0) %>%
  ggplot(aes(x = tripduration)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") +
  geom_density(alpha=.2, fill="#FF6666") + 
  geom_vline(aes(xintercept=mean(tripduration)), color="blue", linetype="dashed", size=1)
```

### Wind Effects on Trip Duration

```{r}
# Average ride time when it's windy
total_data %>%
  filter(AWND > 0) %>%
  summarise(wind_duration_mean = mean(tripduration))

total_data %>% 
  filter(AWND > 0) %>%
  ggplot(aes(x = tripduration)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") +
  geom_density(alpha=.2, fill="#FF6666") + 
  geom_vline(aes(xintercept=mean(tripduration)), color="blue", linetype="dashed", size=1)
```

### Weather Effects on Number of Rides over Average Ride Time

```{r}
# Number of rides over average time sin weather effects
ride_num <- total_data %>%
  filter(tripduration > duration_mean) %>%
  count()
ride_num[1,1]

# Number of rides over average time with rain
rain_num <- total_data %>%
  filter(SNOW == 0, PRCP > 0, tripduration > duration_mean) %>%
  count()
rain_num[1,1]

# Number of rides over average time with snow
snow_num <- total_data %>%
  filter(SNOW > 0, tripduration > duration_mean) %>%
  count()
snow_num[1,1]

# Number of rides over average time with wind
wind_num <- total_data %>%
  filter(AWND > 0, tripduration > duration_mean) %>%
  count()
wind_num[1,1]
```

## Exploratory Data Analysis - Ride History

### Speed of Rider Demographics

```{r}
# Speed of the rider
total_data$speed <- total_data$distance/total_data$tripduration

# Average speed of all riders
all_ride <- total_data %>%
  summarise(average_speed = mean(speed))

# Average speed of young riders
young_ride <- total_data %>%
  filter(age <= 45) %>%
  summarise(young_average = mean(speed))

# Average speed of old riders
old_ride <- total_data %>%
  filter(age >= 65) %>%
  summarise(old_average = mean(speed))

# Average speed of female riders
fem_ride <- total_data %>%
  filter(gender == "Female") %>%
  summarise(female_average = mean(speed))

# Average speed of male riders
male_ride <- total_data %>%
  filter(gender == "Male") %>%
  summarise(male_average = mean(speed))

# Average speed of subscribers
sub_ride <- total_data %>%
  filter(usertype == "Customer") %>%
  summarise(customer_average = mean(speed))

# Average speed of customers
cust_ride <- total_data %>%
  filter(usertype == "Subscriber") %>%
  summarise(subscriber_average = mean(speed))

Reduce(merge, list(all_ride,
                   young_ride,
                   old_ride,
                   fem_ride,
                   male_ride,
                   sub_ride,
                   cust_ride))
```

```{r}
# Scatter Plot of speed by age
total_data %>%
  ggplot(aes(x = age, y = speed, colour = gender)) +
  geom_point(alpha = .4, size = 1.5) +
  scale_colour_manual(name = 'Gender',
                      values = setNames(c('blue','magenta', 'dark green'),
                                        c('Male', 'Female', 'Unknown'))) +
  geom_smooth(method='lm', colour = 'black') +
  labs(title="Average Speed of Riders by Age", x="Speed", y="Age")
```

```{r}
# Boxplot of speed by gender
total_data %>%
  ggplot(aes(x = gender, y = speed, colour = gender)) +
  geom_boxplot(outlier.colour = 'red') +
  scale_colour_manual(name = 'Gender',
                      values = setNames(c('blue','magenta', 'dark green'),
                                        c('Male', 'Female', 'Unknown'))) +
  labs(title="Speed of Riders by Gender", x="Gender", y="Speed")
```

```{r}
# Boxplot of speed by customer type
total_data %>%
  ggplot(aes(x = usertype, y = speed, colour = usertype)) +
  geom_boxplot(outlier.colour = 'red') +
  scale_colour_manual(name = 'User Type',
                      values = setNames(c('purple', 'orange'),
                                        c('Subscriber', 'Customer'))) +
  labs(title="Speed of Riders by Customer Type", x="Customer Type", y="Speed")
```

## Exploratory Data Analysis - Ride History

### Start Locations

```{r}
top_height <- max(total_data$start.station.latitude) - min(total_data$start.station.latitude)
top_width <- max(total_data$start.station.longitude) - min(total_data$start.station.longitude)
top_borders <- c(bottom  = min(total_data$start.station.latitude)  - 0.1 * top_height,
                 top     = max(total_data$start.station.latitude)  + 0.1 * top_height,
                 left    = min(total_data$start.station.longitude) - 0.2 * top_width,
                 right   = max(total_data$start.station.longitude) + 0.2 * top_width)

start <- get_stamenmap(top_borders, zoom = 12, maptype = "toner-lite")
start_map <- ggmap(start, extent = "device", legend = "topright")

start_map + stat_density2d(
aes(x = start.station.longitude, y = start.station.latitude, fill = ..level.., alpha = ..level..,), size = 2, bins = 10, data = total_data, geom = "polygon",  na.rm=TRUE,
) + labs( fill = "Density", title = "Start Location Density") + guides(alpha = F)

```
This graph shows that most bike trips in 2019 start in the center of NYC, with relatively few in the boroughs by comparison.

#### Start Location Preferences - by Day of Week

```{r}
# convert dates to weekdays
total_data$day_of_week = weekdays(total_data$DATE)

start_map +
stat_density2d(
aes(x = start.station.longitude, y = start.station.latitude, fill = ..level.., alpha = ..level..),
size =2, bins = 10, geom = "polygon", data = total_data) + 
guides(alpha = F) + labs(fill = "Density", title = "Start Location Density by Day of Week") +
scale_fill_gradient(low = "yellow", high = "red") +
facet_wrap(~ day_of_week) + 
theme(legend.position = "right")

```
From these charts, we can see that the Saturday and Sunday location densities are slightly more spread than the weekdays, suggesting that the weekend trips are less concentrated in the inner city, albeit still focused in Manhattan. 

#### Start Location Preferences - by User Type

```{r}
start_map +
stat_density2d(
aes(x = start.station.longitude, y = start.station.latitude, fill = ..level.., alpha = ..level..),
size =2, bins = 10, geom = "polygon", data = total_data) + 
guides(alpha = F) + labs(fill = "Density", title = "Start Location Density by User Type") +
scale_fill_gradient(low = "yellow", high = "red") +
facet_wrap(~ usertype) + 
theme(legend.position = "right")

```
This graph shows that there is larger focus on downtown start locations for Subscribers, whereas Customers are spread out along Manhattan and are present in the boroughs as well.

#### Start Location Preferences - by Trip Duration

```{r}
## break down by one standard deviation above and below average of trip duration
ggmap(start) +
    geom_point(data = total_data, mapping = aes(x = start.station.longitude, y = start.station.latitude,
                                        col = tripduration)) +
    scale_color_gradient(low = "yellow", high = "red")
```

### End Location Preferences

```{r}
## before noon and after noon
end <- get_stamenmap(top_borders, zoom = 12, maptype = "toner-lite")
end_map <- ggmap(end, extent = "device", legend = "topright")

end_map + stat_density2d(
aes(x = end.station.longitude, y = end.station.latitude, fill = ..level.., alpha = ..level..),
size = 1, bins = 5, data = total_data,
geom = "polygon"
)
```

#### End Location Preferences - by User Type

```{r}
end_map +
stat_density2d(
aes(x = end.station.longitude, y = end.station.latitude, fill = ..level.., alpha = ..level..),
size =2, bins = 10, geom = "polygon", data = total_data) + 
guides(alpha = F) + labs(fill = "Density", title = "End Location Density by Day of Week") +
scale_fill_gradient(low = "yellow", high = "red") +
facet_wrap(~ day_of_week) + 
theme(legend.position = "right")

```
We see the same pattern breakdown in the end location density as we do the start locations, suggesting that weekends have a more spread out end locations than the weekdays. 

## Asymmetrical Traffic Data 
```{r}
total_rides = count(total_data)
test = total_data
test$start.station.name = as.character(test$start.station.name)
test$end.station.name = as.character(test$end.station.name)
test <- test[test$start.station.name==test$end.station.name, ]
same_station = count(test)

symmetric = same_station / total_rides
asymmetric = 1 - symmetric
```
Only `{r} symmetric`% of rides start and end at the same station, which means that `{r} asymmetric`% of rides are asymmetric traffic.

```{r}
start_popularity = sort(table(total_data$start.station.name), decreasing=TRUE, na.rm=TRUE)
top10 = round(length(unique(total_data$start.station.name, na.rm=TRUE))*0.1)
top_10 = head(start_popularity, top10)
barplot(top_10)
top_starts = as.data.frame(top_10)
top_10rides = sum(top_starts$Freq)

top10_rides = top_10rides / total_rides

```

`{r}top10_rides` of bike rides start come from the top 10% most used station (which are:`{r} top_10`)

```{r}
count_starts = as.data.frame(table(total_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_ends = as.data.frame(table(total_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_starts, count_ends, by.x="station", by.y="station", all.x=TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends

station_flow = na.omit(station_flow)

station_flow %>% mutate(station = fct_reorder(station, net)) %>% ggplot(aes(x=station, y=net)) + geom_bar(stat = "identity")+ geom_hline(yintercept=1, linetype="dashed", color = "red") + labs(x="Stations", y="Total Starts / Total Ends in 2019", title = "Net 2019 Flow of Bikes per Station") + theme(axis.text.x = element_blank())


```
The chart above depicts each station's inflow/outflow of bikes in 2019. Those with a value greater than 1 show that they have a higher rate of bikes starting at their station than ending at their station. These stations would be important to target when thinking about rebalancing bikes, as they overall have more bikes leaving them then coming to them. Similarly, those with the lowest start/end ratios have more bikes ending at their station than leaving, making them prime candidates for moving their surplus to a station in more need.

```{r}
surplus_stations = station_flow[station_flow$net < 0.75,]
deficit_stations = station_flow[station_flow$net > 1.25,]

surplus_stations 
deficit_stations

```
Stations with surplus:  
`{r} surplus_stations `

Stations with deficit: 
`{r} deficit_stations `


