###Average Precipitation by Age
```{r}
###Mean PRCP by Age of Rider
Plotdata <- total_data %>% 
  group_by(age) %>%
  summarise(mean_PRCP_by_age = mean(PRCP),
            mean_duration = mean(tripduration))

ggplot(Plotdata, aes(x = age, y = mean_PRCP_by_age)) + geom_point(alpha =0.9, shape = 18, colour = "blue", size = Plotdata$mean_duration/150) + geom_smooth(colour = "orange") + ggtitle("Average Precipitation During Rides By Rider Age") +
  xlab("Rider Age") + ylab("Mean PRCP")
```
This plot indicates that as rider age increase, the average PRCP during their rides also increases. For ages 40 through 60 however, there is no increse in average PRCP per ride with age. This plot may show that older riders are generally more willing to take rides where the precipitation is greater than younger riders are up to age 65. Data on average PRCP for rides where the rider is over 65-80 years old is sporadic with several outliers.  

###Mean Wind by Age of Rider
```{r}
Plotdata2 <- total_data %>% 
  group_by(age) %>%
   summarise(mean_AWND_by_age = mean(AWND,na.rm = TRUE),
            mean_duration = mean(tripduration)) 

ggplot(Plotdata2, aes(x = age, y = mean_AWND_by_age)) + geom_point(alpha =0.9, shape = 18, colour = "cyan3", size = Plotdata$mean_duration/150) + geom_smooth(colour = "orange") + ggtitle("Average Wind During Rides By Rider Age") +
  xlab("Rider Age") + ylab("Mean Wind")
```
Average wind for rides does not vary much with rider age. The mean wind for rides where the rider is 20 is the same for riders who are 60. However, there is a small dip is average wind for riders aged 25-35 of 0.5 MPH. On more windy days this age group is less willing to use citibike. 

###Mean Snow by Age of Rider
```{r}
Plotdata3 <- total_data %>% 
  group_by(age) %>%
   summarise(mean_SNOW_by_age = mean(SNOW,na.rm = TRUE),
            mean_duration = mean(tripduration)) 

ggplot(Plotdata3, aes(x = age, y = mean_SNOW_by_age)) + geom_point(alpha =0.9, shape = 18, colour = "deepskyblue", size = Plotdata$mean_duration/150) + geom_smooth(colour = "orange") + ggtitle("Average Snow During Rides By Rider Age") +
  xlab("Rider Age") + ylab("Mean Snow")
```
Average snow during rides increases with rider age, although by a small degree of less than 0.01 inches across the 64 ages analyzed. Similar to precipitation data on each age group, the snow data indicates that snow is less of a deterrent to older demographics than it is to younger riders.



#Station locations by most Asymmetric when PRCP >0.3747275
```{r}
big_rain_data <- total_data %>% filter(total_data$PRCP > (0.3747275))
Rain_Locations <- subset(big_rain_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Rain_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_rain_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Rain_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_rain_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]

#Station locations with Net Asymmetric Traffic <0.75, with PRCP >0.3747275
Extra_Bikes_Rain <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Rain <- distinct_data %>% filter(distinct_data$net > 1.25)
Bikes_Rain_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic during High PRCP days
#below 0.75 Asymmetric Traffic
Bikes_Rain_Map + geom_point(data = Extra_Bikes_Rain, aes(x = Extra_Bikes_Rain$long, y = Extra_Bikes_Rain$lat), colour = "deepskyblue") +ggtitle("Most Crowded Stations on High PRCP Days")

##Location of Stations with over 1.25 Asymmetric Traffic during High PRCP days
#More bikes taken than returned in high PRCP
Bikes_Rain_Map + geom_point(data = Lost_Bikes_Rain, aes(x = Lost_Bikes_Rain$long, y = Lost_Bikes_Rain$lat), colour = "deepskyblue") +ggtitle("Most Empty Stations on High PRCP Days")
```
The first map shows the locations of the stations that had significantly more bikes returned than were taken out across all days with more than 1 standard deviation above the mean worth of precipitation, or across all days with greater than 0.3747275 inches of PRCP. Rented bikes at these stations are at most a ratio of 0.75 what the returned biked count during rainy days was.

The second map shows the locations of stations that had many more bikes taken than returned during highly snowy days.Rented bikes are at least 1.25 times greater than returned bikes at all of these stations and wind levels are the same greater than 0.3747275 inches of PRCP as the previous map.

These maps show the same general neighborhoods where stations are overcrowded that they are empty on days with high rain, indicating a risk adjusted system to cover for overcrowded stations on rainy days with empty ones nearby. However, this system may need to be rolled back as there are several station in eastern Brooklyn neighborhoods that are that have no overcrowded stations nearby on days with the same PRCP levels. 

#Station locations by most Asymmetric when SNOW > 0.21624501
```{r}
big_snow_data <- total_data %>% filter(total_data$SNOW > (0.21624501))
Snow_Locations <- subset(big_snow_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Snow_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_snow_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Snow_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_snow_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]


Extra_Bikes_Snow <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Snow <- distinct_data %>% filter(distinct_data$net > 1.25)
Bikes_Snow_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic during High Snow days
#below 0.75 Asymmetric Traffic
Bikes_Snow_Map + geom_point(data = Extra_Bikes_Snow, aes(x = Extra_Bikes_Snow$long, y = Extra_Bikes_Snow$lat), colour = "cyan") +ggtitle("Most Crowded Stations on High Snow Days")

##Location of Stations with over 1.25 Asymmetric Traffic during High Snow days
#More bikes taken than returned in high Snow
Bikes_Snow_Map + geom_point(data = Lost_Bikes_Snow, aes(x = Lost_Bikes_Snow$long, y = Lost_Bikes_Snow$lat), colour = "cyan") +ggtitle("Most Empty Stations on High Snow Days")
```
The first map shows the locations of the stations that had significantly more bikes returned than were taken out across all days with more than 1 standard deviation above the mean worth of snow, or across all days with greater than 0.21624501 inches of snow. Rented bikes at these stations are at most a ratio of 0.75 what the returned biked count during snowy days was.

The second map shows the locations of stations that had many more bikes taken than returned during highly snowy days.Rented bikes are at least 1.25 times greater than returned bikes at all of these stations and wind levels are the same greater than 0.21624501 inches of snow as the previous map.

The same trend can be observed in these maps as in the high PRCP, high asymmetry maps above. Most of the overcrowded stations are surrounded by emptier stations to cover for customers trying to return bikes. However, these maps also follow a trend of the deep areas in east Brooklyn have many empty stations. The reason for many emptier staions being loccated in these areas during both high PRCP and high snow days may be the remote location of these stations in relation to major hubs of the city. Riders who are initially willing to ride through these types of streets and these distances at earlier points of the day may not be willing to return via citibike to east Brooklyn. These streets are ore residential and less likely to be as well kept and prioritized by the city. In addition, the later in the day the darker it may be and the more snowy or wet the street may be. 

#Station locations by most Asymmetric when AWND > 6.383933
```{r}
big_wind_data <- total_data %>% filter(total_data$AWND > (6.383933))
Wind_Locations <- subset(big_wind_data, select=c("start.station.longitude", "start.station.latitude", "start.station.name"))
names(Wind_Locations) = c("long", "lat", "station")
count_starts = as.data.frame(table(big_wind_data$start.station.name), na.rm=TRUE)
names(count_starts) = c("station", "starts")
count_begins = as.data.frame(merge(count_starts,Wind_Locations, by.x = "station", by.y = "station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
count_ends = as.data.frame(table(big_wind_data$end.station.name), na.rm=TRUE)
names(count_ends) = c("station", "ends")
station_flow = as.data.frame(merge(count_begins, count_ends, by.x="station", by.y="station", all.x=TRUE, all.y = TRUE, na.rm=TRUE))
station_flow$net = station_flow$starts / station_flow$ends
station_flow = na.omit(station_flow)
distinct_data <- station_flow[!(duplicated(station_flow$station)), ]


Extra_Bikes_Wind <- distinct_data %>% filter(distinct_data$net < 0.75)
Lost_Bikes_Wind <- distinct_data %>% filter(distinct_data$net > 1.25)
Bikes_Wind_Map <- ggmap(start, extent = "device", legend = "topright")

##Location of Stations with Significantly Positive Asymmetric Traffic during high wind days
#below 0.75 Asymmetric Traffic
Bikes_Wind_Map + geom_point(data = Extra_Bikes_Wind, aes(x = Extra_Bikes_Wind$long, y = Extra_Bikes_Wind$lat), colour = "darkorchid1")+ggtitle("Most Crowded Stations on High Wind Days")

##Location of Stations with over 1.25 Asymmetric Traffic during High Snow days
#More bikes taken than returned in high wind
Bikes_Wind_Map + geom_point(data = Lost_Bikes_Wind, aes(x = Lost_Bikes_Wind$long, y = Lost_Bikes_Wind$lat), colour = "darkorchid1") + ggtitle("Most Empty Stations on High Wind Days")
```
The first map shows the locations of the stations that had significantly more bikes returned than were taken out across all days with more than 1 standard deviation above the mean worth of wind speed, or across all days with greater than 6.383933 miles per hour. Rented bikes at these stations are at most a ratio of 0.75 what the returned biked count during windy days was. It shows where positive asymmetric traffic is concentrated on days with a high level of wind. Most of these stations are located in Brooklyn and Queens, rather than Manhattan. It is clear than most of the stations that may face an issue with overcrowding on windy days are in the Myrtle area of Brooklyn and western Queens stations. 

The second map shows the locations of stations that had many more bikes taken than returned during highly windy days.Rented bikes are at least 1.25 times greater than returned bikes at all of these stations and wind levels are the same greater than 6.383933 miles per hour as the previous map. This map shows a few more stations popping up in the Upper East and Upper West Sides of Manhattan as well as a shift even further west of the western stations in Queens observed in the previous map. Most significantly different in this map is the increase in the number of stations in same Brooklyn neighborhood. Since there is already a large number of close stations that are overcrowded close by in both Brooklyn and Queens neighborhoods, it could be concluded that keeping these more empty stations avaliable close are for customers who run into an overcrowded station.



